Rust Bug Fix Patterns

### 2022-09-12 3_results.tex
added some descriptions for results and reorganized files

\documentclass{article}
\usepackage{listings, listings-rust}
\lstset{escapeinside={<@}{@>}}
\usepackage{minted}
\usepackage{tabularx}
\usepackage{array}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[utf8]{inputenc}
\usepackage{multirow}

\title{Rust Bug Fix Patterns}
\author{Moh Robati}

\begin{document}

\maketitle

\section{Introduction}

Producing bugs has been an unavoidable part of computer programming which happens unintentionally. Researchers have been trying to help software engineers to be more aware of the bug producing practices in development and how to step towards creating bug-free software. One useful step towards that goal would be studying common bug patterns (and their corresponding fixes) that occur across multiple software systems. Research efforts for categorizing common bug patterns date back to 1975 where Edres tried to categorize various bug patterns which occurred in operating system program.

In this work, we empirically analyze the most popular Rust projects in GitHub, and discover the common bug patterns lied within them. We believe these insights are important for various reasons, primarily for developing program repair tools. Automated program repair tools try to locate bug
locations in the source code (they include a fault localization module) and then fix them, possibly without human intervention, in a manner such that the modified program meets the desired specifications. A powerful automated program repair system has many useful applications and can significantly simplify the process of debugging, resulting in a reduction in software development cost. The set of common bug fix patterns can be a useful input to an APR system as it directly provides information about fixing similar buggy structures, and that can significantly reduce the search space for patch generation.

Similarly, if we are not observing the desired behaviour from our program, that would indicate that there lies a fault within our program. Fault localization is the process of automatically finding the faulty statement(s) in our program. If a fault localization tool knows about the common bug patterns that can appear in the underlying programming language, it can prioritize the code segment candidates based on their potentials of producing bugs. That would result in a more robust tool as it leverages from historical data.

Moreover, knowing which code patterns in a programming language are more susceptible to produce bugs can help programmers to be more cautious while working with similar code structures. This knowledge can make the developers avoid certain mistakes during production which ultimately would result in reducing the frequent bug patterns within the program. As Monperrus describes, the less frequent bug patterns we keep in our software, the less cost we would have to pay for software maintenance.

To discuss about a less talked about use case, recently, researchers have been trying to apply deep learning techniques on programs to learn about the semantics of the code or training models which can modify programs towards a specific goal (e.g. bug fix). One challenge is that usually learning tools get fixed size vectors as input and map that to an output. Unlike image data, it is challenging to compress programs into a vector without any loss in semantic information. Current approaches try to define the vector dimensions in a way that accounts for all possible abstract syntax tree forms. That results in a large sparse space for the inputs. Assuming that the deep learning model we are trying to develop is associated with bug fixing tasks, then knowledge about common bug fix patterns can help with pruning input space towards containing less dimensions. That can help with creating more efficient models.

There have been many studies on detecting and categorizing bug fix patterns in general, or for a specific programming language. We are more interested in the latter, since
our study is focused on pervasive bug fix patterns in Rust. \\
Yang et al. proposed a mining approach to detect Python bug fix patterns via studying
fine-grained fixing code changes. In their research, they also examined what portion of the bugs
could be fixed using automated bug fixing approaches. Moreover, they evaluated the fix patterns in
the wild and realized that 37\% of the buggy codes could be matched by the fix patterns they had
found. \\
Hanam et al. conducted a similar research but for finding Javascript most pervasive bug fix
patterns. They presented a large-scale study of the most common bug fix patterns by mining 105K
commits from 134 server-side JavaScript projects. They used DBSCAN clustering algorithm
and divided bug fixing change types into 219 clusters, from which they extracted 13 pervasive
cross-project bug fix patterns. \\
In this work, we study common bug patterns that appear in Rust projects. Rust is a relatively new language that allows programmers to write programs that have low-level
control over resources, as well as high-level safety guarantees. Originally, Rust was a side
project developed by Graydon Hoare, a Mozilla employee. The company started to fund this project
in 2009, since they intended to use Rust in some of the low-level Firefox components.
What distinguishes Rust from the other systems programming languages, e.g. C/C++, is the
ownership framework that has been incorporated inside the language. The ownership framework keeps the track of the references to different memory locations, and will free them once they
go out of scope. If the program passes the compiler checks, it automatically guarantees several
safety problems e.g. dangling references, or double-free problem, that had to be taken care of
manually in languages like C/C++. Moreover, not only the ownership concept removes the
need for a garbage collector in the runtime environment, it does not require the programmer to
explicitly do certain safety tasks, like freeing references.
Due to the memory and thread safety guarantees that Rust provides, many software companies
adopted Rust in production. For example, Dropbox decided to migrate to Rust for developing their
storage system, since they were not completely satisfied with the features provided by AWS’s S3 or
Golang.

The main module which governs over the ownership rules is called Borrow Checker. Because of its importance in Rust, we have decided to put discovered patterns into two groups; (1) non-borrow-checker related patterns (General Patterns), and (2) borrow-checker related patterns (BC-Related Patterns). This distinction was really helpful as it made our tool discover clearer results.

We present our discovered insights of our empirical study in the following four research
questions (RQs):

\textbf{RQ1. How successful is our data modelling approach in capturing all the necessary data?}

\textbf{RQ2. What are the most common AST structures in the bug fixing changes?}

\textbf{RQ3. What are the general fix patterns in Rust?}

\textbf{RQ4. Which fix patterns are related to Borrow Checker in Rust?}

\section{Pervasive Bug Patterns}
\subsection{Attributes}

@@

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
-- return Poll::Ready(ReadyEvent { ready, tick });
++ return Poll::Ready(ReadyEvent { tick, ready });
\end{lstlisting}

@@

@@

pub struct Manifest {
  -- pub features: Vec<String>,
  ++ pub features: HashSet<String>,
}

@@

@@

@@

@@

@@

@@

@@

@@

\subsubsection{B3. Dropping \& from func decl and adding * to match statement}

@@

\subsubsection{B4. Adding *}

@@

@@

@@

@@

@@

\subsection{Clone}
\subsubsection{B11. Dropping clone and adding dereference}

@@

@@

@@

\multirow{12}{*}{General}        
& \multirow{2}{*}{Attributes}    & G1  & Adding Attributes to Structs                                        \\
                                 &                                & G2  & Adding Attributes to Struct Fields                                  \\\cline{2-3}
                                 & \multirow{3}{*}{Struct}        & G3  & Adding/Removing a Struct field                                      \\
                                 &                                & G4  & Adding/Removing pub to/from struct field                            \\
                                 &                                & G5  & Different fields order when initializinng Struct                       \\\cline{2-3}
                                 & Option                         & G6  & Changing field type T to Option                                          \\\cline{2-3}
                                 & \multirow{4}{*}{Types}         & G7  & Changing a Struct field type                                        \\
                                 &                                & G8  & Changing a generic type                                             \\
                                 &                                & G9  & Change in enum variant value type                                   \\
                                 &                                & G10 & Changing the type inside a tuple                                         \\\cline{2-3}
                                 & \multirow{2}{*}{Traits}        & G11 & TraitBound Change                                                   \\
                                 &                                & G12 & Implementing a different trait for a struct.                            \\\cline{1-3}
\multirow{13}{*}{Borrow Checker} & \multirow{4}{*}{Ref and Deref} & B1  & Adding \&                                                           \\
                                 &                                & B2  & Removing \&                                                         \\
                                 &                                & B3  & Dropping \& from function declaration and dereferencing match \\
                                 &                                & B4  & Adding *                                                            \\\cline{2-3}
                                 & \multirow{2}{*}{Mut}           & B5  & Adding mut                                                          \\
                                 &                                & B6  & Removing mut                                                        \\\cline{2-3}
                                 & Vector                         & B7  & Vector reference to slice                                           \\\cline{2-3}
                                 & \multirow{3}{*}{Lifetime}      & B8  & Adding generic Lifetime                                             \\
                                 &                                & B9  & Dropping static lifetime                                            \\
                                 &                                & B10 & Dropping Lifetime in field definition                               \\\cline{2-3}
                                 & \multirow{3}{*}{Clone}         & B11 & Dropping clone and adding dereference                               \\
                                 &                                & B12 & Dropping clone                                                      \\
                                 &                                & B13 & Dropping clone and adding reference

@@

\begin{algorithm}
\caption{An algorithm with caption}\label{alg:cap}
\hspace*{2mm} \textbf{Input:} $R$ (target repositories)  \\
\hspace*{2mm} \textbf{Output:} $D_g$ (General code changes) \\
\hspace*{2mm} \textbf{Output:} $D_b$ (BC-related code changes)
\begin{algorithmic}
\State $D_g \leftarrow \phi$
\State $D_b \leftarrow \phi$
\For{$r \in R$}
    \For{$c \in \textsc{ExtractCommits}(r)$}
        \If{$c.msg$ contains bug fixing related keywords}
            \For{$\{f_b, f_a\} \in \textsc{GetModifiedRustFiles}(c)$}
                \For{$e \in \textsc{ASTDiff}(\textsc{Parse}(f_b), \textsc{Parse}(f_a))$}
                    \State $DP \leftarrow \textsc{GetDataPoint}(e)$
                    \State $DP \leftarrow c.hash \cup f_a.name \cup e.scope \cup DP $
                    \If{$\textsc{IsBCRelated}(e)$}
                        \State $DP \leftarrow \textsc{GetBCKeyword}(e) \cup DP $
                        \State $D_b \leftarrow D_b \cup DP$
                    \Else
                        \State $D_g \leftarrow D_g \cup DP$
                    \EndIf
                \EndFor
            \EndFor
        \EndIf
    \EndFor
\EndFor
\end{algorithmic}
\end{algorithm}

\end{document}

>>>>

\subsection{Attributes}

@@

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
-- return Poll::Ready(ReadyEvent { ready, tick });
++ return Poll::Ready(ReadyEvent { tick, ready });

\end{lstlisting} \\

\noindent \textbf{Description:} Remove... ???

@@

\noindent \textbf{Description:} Rust is a null safe language, meaning that object references cannot assume null values. If we want void values for our object, we would have to use Option type. Option is essentially an enum with two variants, Some which indicates that our object has an value, and None shows the absence of any value. In this pattern, the developer changes the type of a struct field to an Option of its type. This makes it possible for the developer to modify the program in a way that it covers the scenarios in which the the struct field has no value.

For instance, in the Rust project swc-project/swc (A fast TypeScript/JavaScript compiler), the developer needed to modify the CSS parser source code in a way that it accounts for empty @media queries \footnote{\href{https://github.com/swc-project/swc/commit/75a14f98b7370226115ee24eec6eb8c802bd4837}}, so, as shown in the code snippet above, they changed type MediaQueryList to Option<MediaQueryList>, and modified the other parts of the source code, accordingly.

@@

\begin{lstlisting}[language=Rust, style=colouredRust, label={l3}]
pub struct Manifest {
  -- pub features: Vec<String>,
  ++ pub features: HashSet<String>,
}

@@

\noindent \textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. These reasons can be a step for implementing a new feature or rectifying the behaviour of the program. Also it can happen for refactoring purposes or performance enhancement.

@@

\noindent \textbf{Description:} This a change of the generic (parametric) type of a struct field. Similar to G7, this change can also happen for the mentioned reasons.

@@

\noindent \textbf{Description:} In Rust, we can specify the type of the value that an enum variant can store. Using this feature, we can avoid using a struct alongside our enum variants to specify the value type. This change pattern is associated with a change in enum variant types.

@@

\noindent \textbf{Description:} In this pattern, an extra type is added to an existing type, making it a tuple of multiple types. Alternatively, the change might be dropping types from the tuple and making it a tuple with fewer number of arguments.

@@

\noindent \textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This change pattern related to modification in trait bounds. This can happen both for bug fix reasons or just refactoring purposes.

@@

\noindent \textbf{Description:} Remove... ???

@@

\noindent \textbf{Description:} This change happens when the developer needs to modify the program so it reaches the desired behaviour. The modification, however, requires access to an object whose ownership has moved to another scope and never returned. That is why we must use borrowing to preserve ownership within our scope, after the function execution finishes.

To give an example, in starship/starship after one of the commits \footnote{\href{https://github.com/starship/starship/commit/56d475578ea508631275772127f49a6949fea6b0}{git_branches.rs}}, the developer borrows the value of $repo\_dir$ to keep the ownership of the object in the current scope. Now this object can then be used in another function invocation ($remove\_dir\_all(repo\_dir)$) which fixes a bug.

@@

\noindent \textbf{Description:} In the pattern, an unnecessary borrowing is removed. Clippy can detect the needless borrows, and also can automatically remove them. This does not change the behaviour of the program, however, it does improve the performance, as the borrow checker need not to return the ownership after the execution of the function.

\subsubsection{B3. Dropping & from func decl and adding * to match statement}

@@

\noindent \textbf{Description:} ???

\subsubsection{B4. Modification in dereferencing a string prior to calling .to_string()}

@@

\noindent \textbf{Description:} This is a change that in all the instances has been detected by Clippy. The change applies on an object prior to calling $to\_string()$ on that object. The change might add or remove a dereference, as shown above. In the addition case, the reason behind the change is that the object a reference type of $T$, while $T$ directly implements $to\_string()$. It is faster and more performance efficient to dereference the object and then call $to\_string()$ on it. The deletion case indicates a redundant dereference, thus it is removed.

@@

\noindent \textbf{Description:} This change happens when the programmer needs to mutate a variable which is immutable, hence they need to add the mut keyword. In the observed instances, this pattern was not seen in refactoring changes, the modification happened alongside a change in program behaviour.

@@

\noindent \textbf{Description:} Dropping the mut keyword before a variable, as we do not mutate the variable. This redundant mutability is reported by Clippy and can also be removed through this tool.

@@

\noindent \textbf{Description:} ???

@@

\noindent \textbf{Description:} ???

\subsection{Clone}
\subsubsection{B11. Dropping clone (and adding dereference)}

@@

\noindent \textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy detects whether a variable clone is necessary and could we have used the variable itself or not. This change does not modify the program behaviour, however it can boost the performance significantly if we repeatedly cloning large variables.

@@

\noindent \textbf{Description:} Unlike the previous pattern, this pattern is not detected by Clippy. The change happens when the programmers realizes that the cloning of a variable is unnecessary, however, we also want to preserve the ownership of the variable within the current scope. That is why in this pattern a clone is turned into a borrowing.

Similar to the previous pattern, this does not change the program behaviour and is merely for performance purposes. It was the case in one of the commits of the project tauri-apps/tauri, where the clone to borrowing significantly optimized CPU usage. They discussed that the cloning of some types such as HashMap was expensive.

@@

\multirow{12}{*}{General}        
& \multirow{2}{*}{Attributes}    & G1  & Adding Attributes to Structs                                        \\
                                 &                                & G2  & Adding Attributes to Struct Fields                                  \\\cline{2-3}
                                 & \multirow{3}{*}{Struct}        & G3  & Adding/Removing a Struct field                                      \\
                                 &                                & G4  & Adding/Removing pub to/from struct field                            \\
                                 &                                & G5  & Different fields order when initializinng Struct                       \\\cline{2-3}
                                 & Option                         & G6  & Changing field type T to Option                                          \\\cline{2-3}
                                 & \multirow{4}{*}{Types}         & G7  & Changing a Struct field type                                        \\
                                 &                                & G8  & Changing a generic type                                             \\
                                 &                                & G9  & Change in enum variant value type                                   \\
                                 &                                & G10 & Changing the type inside a tuple                                         \\\cline{2-3}
                                 & \multirow{2}{*}{Traits}        & G11 & TraitBound Change                                                   \\
                                 &                                & G12 & Implementing a different trait for a struct.                            \\\cline{1-3}
\multirow{13}{*}{Borrow Checker} & \multirow{4}{*}{Ref and Deref} & B1  & Adding \&                                                           \\
                                 &                                & B2  & Removing \&                                                         \\
                                 &                                & B3  & Dropping \& from function declaration and dereferencing match \\
                                 &                                & B4  & Adding *                                                            \\\cline{2-3}
                                 & \multirow{2}{*}{Mut}           & B5  & Adding mut                                                          \\
                                 &                                & B6  & Removing mut                                                        \\\cline{2-3}
                                 & Vector                         & B7  & Vector reference to slice                                           \\\cline{2-3}
                                 & \multirow{3}{*}{Lifetime}      & B8  & Adding generic Lifetime                                             \\
                                 &                                & B9  & Dropping static lifetime                                            \\
                                 &                                & B10 & Dropping Lifetime in field definition                               \\\cline{2-3}
                                 & \multirow{3}{*}{Clone}         & B11 & Dropping clone and adding dereference                               \\
                                 &                                & B12 & Dropping clone                                                      \\
                                 &                                & B13 & Dropping clone and adding reference  & \hline

@@
